Kinematics of the Palomar 5 stellar stream from RR Lyrae stars

### 2019-08-21 paper/main.tex
shuffling

\begin{abstract}
\todo{This is just placeholder text!}
% Context
Stellar streams are important.
The Palomar 5 stellar stream is a prototypical example: It connects back to the progenitor cluster.
RR Lyrae-type (RRL) stars are useful because they are standard candles and their distances can be inferred with small errors ($\sim7\%$)
% Aims
While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
% Methods
Here we use a large catalog of RRL stars released as a part of Gaia DR2 to search for RRL stars in the Pal 5 stream.
By building a probabilistic model of the cluster and stream in proper motion and distance, we find XX RRL consistent with being members of the stream.
% Results
The number of Pal 5 RRL is therefore XX times larger in the tails than in the remnant cluster: This changes our interpretation of ...blah.
% Conclusions
Stuff.
\end{abstract}

@@

To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G~21$ for the typical colors of RRLs).
Periods and amplitudes in all five bands are available for all stars in this catalog.
We have chosen to use this catalog instead of the recently released \Gaia~RRL catalogs \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018} as these are subject to higher incompleteness in this particular area of the sky due to the \Gaia\ scanning law. Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.

@@

To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components ($\approx 5~\kms$ at the distance of Pal 5) and $0.2~\kpc$ in distance.
However, we allow the mean to vary independently in each component as a function of $\phi_1$, a longitude coordinate (along the stream) as defined in \citet{Bonaca:2019}.
Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
The mean of our stream component, $\bs{x}$, is therefore given by
\begin{equation}
    \bs{x} = \begin{pmatrix}
        \mu_\alpha^* + b_\alpha \, \phi_1 + c_\alpha \, \phi_1^2\\
        \mu_\delta^* + b_\delta \, \phi_1 + c_\delta \, \phi_1^2\\
        d^* + b_d \, \phi_1 + c_d \, \phi_1 \label{eq:meanpoly}
    \end{pmatrix}
\end{equation}
where $\bs{\theta} = (\mu_\alpha^*, \mu_\delta^*, d^*, b_\alpha, b_\delta, b_d, c_\alpha, c_\delta, c_d)$ are free parameters.

@@

\todo{How many we find, what their types are}

@@

\todo{Summarize mean dist and pm from RRLs, and trend parameters}

\todo{Make a table with summary information}

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}\label{fig:PA_diagram}
\end{center}
\end{figure}

\citet{Sesar2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations. They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails. At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%. We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner. Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.

\subsection{Implications for stellar population}

\todo{Constraints on initial mass of Pal 5}

\todo{CM:} Inferred luminosity based on the number of RRLs is XX. Compare to inferred luminosity based on the $M_V-\FeH$ relation.

\subsection{Cluster distance and stream track}

>>>>

\begin{abstract}
% Context
Thin stellar streams, formed from the tidal disruption of globular clusters, are important gravitational tools, sensitive to both global and small-scale properties of dark matter.
The Palomar 5 stellar stream (Pal 5) is an exemplar stream within the Milky Way: Its $\approx 20\degr$ stream connects back to the progenitor cluster, and it has been used to study the shape and total mass of, along with substructure in, the dark matter distribution of the Galaxy.
However, ...most details of the stream are not fully explained.
% Aims
...we want to measure distance and kinematic tracks along the stream to inform models.
Here we use a large catalog of RR Lyrae-type stars (RRLs) with astrometric data from the \Gaia\ mission to probabilistically identify RRLs in the Pal 5 stream.
RRLs are useful because they are intrinsically-luminous standard candles and their distances can be inferred with small relative precision ($\sim3\%$).
% While the Pal 5 cluster is known to host a small number (5--7) of RRL stars, only two RRL stars have since been found in the long tidal tails that form the Pal 5 stream.
% Methods
% Results
By building a probabilistic model of the cluster and stream in proper motion and distance, we find XX RRL consistent with being members of the stream.
The number of Pal 5 RRL is therefore XX times larger in the tails than in the remnant cluster: This changes our interpretation of ...blah.
% Conclusions
Stuff.
\end{abstract}

@@

To search for RRLs in the vicinity of Pal 5, we use a catalog of RRL stars from the PS1 survey that spans $\sim$3/4 of the sky (all sky north of declination $\delta > -30\degr$) and contains 61\,795 RRLs identified from their variability \citep{Sesar:2017b}.
In detail, the \emph{bona fide} RRLs in the catalog were classified using multi-epoch data in the $grizy$ photometric bands, with a typical number of 12 observations per filter, and has a limiting magnitude $r\sim21$ (corresponding to $G \sim 21$ for the typical colors of RRLs).
Periods and amplitudes in all five bands are available for all stars in this catalog.
We have chosen to use this catalog instead of the recently released \Gaia~RRL catalogs \citep[VariClassifier and Specific Object Studies][]{Holl2018, Rimoldini2018, Clementini2018} as these are subject to higher incompleteness in this particular area of the sky due to the \Gaia\ scanning law. Nevertheless, the vast majority (61\,625) of PS1's \emph{bona fide} RRLs have astrometric information available in \Gaia~\DR{2}.

@@

To represent the Pal 5 stream, we use a single Gaussian component with a fixed dispersion of $0.05~\masyr$ for the proper motion components ($\approx 5~\kms$ at the distance of Pal 5) and $0.2~\kpc$ in distance.
However, we allow the mean to vary independently in each component as a function of $\phi_1$, a longitude coordinate (along the stream) as defined in \citet{Bonaca:2019}.
Based on the Pal 5 stream models from \citet{Bonaca:2019}, over the range of $\phi_1 \in (-20^\circ, 15^\circ)$, the mean stream trends in $\mu_\alpha$, $\mu_\delta$, and distance can be well-approximated (within our observational uncertainties) by a quadratic function in $\phi_1$.
The mean of our stream component, $\bs{x}$, is therefore given by
\begin{equation}
    \bs{x} = \begin{pmatrix}
        \mu_\alpha^* + b_\alpha \, \phi_1 + c_\alpha \, \phi_1^2\\
        \mu_\delta^* + b_\delta \, \phi_1 + c_\delta \, \phi_1^2\\
        d^* + b_d \, \phi_1 + c_d \, \phi_1 \label{eq:meanpoly}
    \end{pmatrix}
\end{equation}
where $\bs{\theta} = (\mu_\alpha^*, \mu_\delta^*, d^*, b_\alpha, b_\delta, b_d, c_\alpha, c_\delta, c_d)$ are free parameters.

@@

We find a total of 24 RR Lyrae stars with membership probability $> 0.5$ that lie within $1\degr$ of the stream track \citep[again using the track from][]{Bonaca:2019}.
Of those, 13 are classified as type \typeab\ and 11 are classified as type \typec\ RRLs.
We find a total of 10 RRLs within the estimated Jacobi radius of the cluster ($\approx 12'$), meaning that 15 RRLs are spread between the leading and trailing tails.
\figurename~\ref{fig:members} shows the sky positions of all RRL with membership probability $>0.01$ over-plotted on the density of filtered main sequence stars from \citet{Bonaca:2019}.
We note that there are a few RRLs outside of the main stream track that appear to be at a consistent distance with consistent proper motions (e.g., the two stars near $230\degr, 0\degr$) that are not included in the tally above.
No theoretical models of Pal 5 have predicted tidal debris at such locations, so follow-up spectroscopy to confirm or rule out their origin would be very informative.

@@

We find a mean Pal 5 cluster distance of $20.9 \pm 0.2~\kpc$ using the PS1 RRLs and the membership model described in \sectionname~\ref{sec:membership}.
This is significantly closer than past measurements of the cluster distance that have used \todo{summarize past distance measurements and why this could be so different...}.
We also detect a distance trend from the closer leading tail (at $\approx 19~\kpc$) to the more distant trailing tail (at $\approx 22~\kpc$).
While the proper motion in right ascension appears consistent with constant or only a mild gradient over the extent of the stream, we see a significant gradient in the declination component, with a difference of about $\approx 1~\masyr$ from trailing to leading tail.
% \todo{Make a table with summary information?}

\begin{figure*}[t]
\begin{center}
\includegraphics[width=\textwidth]{rrls_PA.pdf}
\caption{Amplitude ($g$-band) versus Period for RRLs in the Pal~5 field.}
\label{fig:PA_diagram}
\end{center}
\end{figure*}

\citet{Sesar:2017b} provide an estimate of their catalogue's completeness as a function of distance, at high galactic latitude, based on simulations. They estimate the mean completeness to be 92\% for \rrab~and 79\% for \rrc, up to a distance of 40~kpc, well beyond the distance of Pal~5 and its tails. At the average ratio of 3 \rrc~stars per every 10 \rrab~\citep{Layden1995}, this represents a mean completeness of 89\%. We can also produce an estimate per line of sight using the methodology described in \citet{Rybizki2018}, which requires two independent catalogues to assess the completeness of both in a probabilistic manner. Using the \Gaia~DR2 VariClassifier \citet{Holl2018,Rimoldini2018} and Specific Objects Study \citep{Clementini2018} catalogues, we find a median completeness of 92\% with a standard deviation of 16\%  with no evident spatial trend across the field of Figure~\ref{fig:members}.

\section{Discussion}

@@

\section{Conclusions} \label{sec:conclusions}

\todo{what did we do}
